---
title: "power analysis - binomial GLMM - intercept-only"
author: "Christoph Völter"
date: "31/05/2020"
output: 
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(cowplot)

load("power_sim_disjsyll_order_as_factor_1000iter_intercept_only.RData")
```

## Generate data

```{r echo=FALSE, include=FALSE}

set.seed(1)
n.subject <- 16 # number subjects
n.per.subject <- 12 # observations per subject
n.per.condition <- 12 # observations per subject and condition
subj.id <- as.factor(paste("subj", 1:n.subject, sep = "."))
control.per <- 1 / 3 # performance in control condition
test.per <- c(0.55, 0.65) # performance in test condition

start.data <- data.frame(subj.id)
# duplicate rows according to the number obs. per subject:
start.data <- start.data[rep(x = 1:nrow(start.data), times = n.per.subject), ]
start.data <- as.data.frame(start.data)
names(start.data) <- "subj.id"

# add condition and trial number
start.data <- data.frame(expand.grid(subj.id = subj.id, condition = c("control", "test"), trial = c(1:n.per.condition)))

# add order
start.data$order <- as.factor(rep(x = c("test_first", "control_first"), times = n.subject / 2))[as.numeric(start.data$subj.id)]

start.data$order2 <- ifelse((start.data$order == "test_first" & start.data$condition == "test") |
  (start.data$order == "control_first" & start.data$condition == "control"), 1, 2)


# add demographics  (if it should be generated in each loop)
chimp.demo <- read.csv("data/chimp_demographics.csv")
start.data$subject_name <- as.factor(rep(x = chimp.demo$Subject, times = n.per.subject))[as.numeric(start.data$subj.id)]
start.data$age <- as.numeric(rep(x = chimp.demo$Age, times = n.per.subject))[as.numeric(start.data$subj.id)]
start.data$sex <- as.factor(rep(x = chimp.demo$Sex, times = n.per.subject))[as.numeric(start.data$subj.id)]


# z-transformation of covariates
start.data$z.age <- as.vector(scale(start.data$age))
start.data$z.trial <- as.vector(scale(start.data$trial))
start.data$z.order <- as.vector(scale(as.numeric(start.data$order2)))

# dummy code factors
start.data$condition.dummy <- as.numeric(start.data$condition == levels(start.data$condition)[2])
start.data$order.dummy <- as.numeric(start.data$order == levels(start.data$order)[2])

# center condition for random slopes:
start.data$condition.c <- as.numeric(start.data$condition) - mean(as.numeric(start.data$condition))
start.data$order.c <- as.numeric(start.data$order) - mean(as.numeric(start.data$order))

# checks:
# does each subject have only one sex and age?
xx <- table(start.data$subj.id, start.data$sex)
range(apply(X = xx > 0, MARGIN = 1, sum)) # should be 1 and 1

xx <- table(start.data$subj.id, start.data$age)
range(apply(X = xx > 0, MARGIN = 1, sum)) # should be 1 and 1

xx <- table(start.data$subj.id, start.data$condition)
range(apply(X = xx > 0, MARGIN = 1, sum))

xx <- table(start.data$subj.id, start.data$trial)
range(apply(X = xx > 0, MARGIN = 1, sum))

xx <- table(start.data$condition, start.data$order)
range(apply(X = xx > 0, MARGIN = 1, sum))

xx <- table(start.data$subj.id, start.data$order)
range(apply(X = xx > 0, MARGIN = 1, sum))
```



## Simulation

```{r eval=FALSE, include=FALSE}
n.simus <- 1000 # small number for testing
r.effects <- c(0.175, 0.35, 0.7) # random effects to be simulated
# with the intercept being -0.6931472 (qlogis(1/3)) we could make the following
# guesses for the random intercept:
#- 0.175: tiny  random intercepts effect
#- 0.35: moderate random intercepts effect
#- 0.7: strong random intercepts effect
#- 1.4: very strong random intercepts effect

r.slope.trial <- 0.1

# create object to store the simulation parameters and results:
all.res <- data.frame(expand.grid(
  n.per.subject = n.per.subject, r.effect = r.effects,
  r.slope.trial = r.slope.trial,
  test.per = test.per,
  simu = 1:n.simus
))
all.res$icpt <- NA
all.res$icpt.p <- NA
all.res$re.sd <- NA
all.res$warns.full <- NA
all.res$warns.null <- NA


# create data frame with design:
## done above

# load packages needed:
library(lme4)
# Loading required package: Matrix
library(kyotil) # we want to store info about convergence issues

# define control structure to make convergence more likely:
contr <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 10000))

xdata <- start.data %>%
  filter(condition == "test")

m.mat <- model.matrix(object = ~z.trial, data = xdata) # create model martix

# run simulation
for (i in 1:nrow(all.res)) {
  set.seed(i) # allows to later replicate individual simulations

  coefs <- c(
    "(Intercept)" = log(all.res[i, "test.per"] / (1 - all.res[i, "test.per"])),
    "z.trial" = 0
  )

  LP <- m.mat[, names(coefs)] %*% coefs # LP wrt fixed effects

  # add random effect to linear predictor:
  LP <- LP + rnorm(n = n.subject, sd = all.res[i, "r.effect"])[as.numeric(xdata$subj.id)] +
    rnorm(n = n.subject, sd = all.res[i, "r.slope.trial"])[as.numeric(xdata$subj.id)] * xdata$z.trial

  # generate response:
  xdata$correct <- rbinom(n = nrow(xdata), size = 1, prob = exp(LP) / (1 + exp(LP)))

  # fit full model:
  full <- keepWarnings(glmer(correct ~ 1 + (1 + z.trial || subj.id),
    data = xdata, family = binomial, control = contr
  ))

  # store results:
  all.res[i, c("icpt")] <- fixef(full$value)
  all.res[i, "icpt.p"] <- summary(full$value)$coefficients["(Intercept)", "Pr(>|z|)"]
  all.res[i, "re.sd"] <- as.data.frame(summary(full$value)$varcor)[1, "sdcor"]
  all.res[i, "warns.full"] <- nchar(paste(full$warnings, collapse = ""))
}


#save.image("power_sim_disjsyll_order_as_factor_1000iter_intercept_only.RData")
```



## Evaluation of results 

* number of warning per combinations of random effects (out of 1000 models per cell)  
Full model:  
```{r echo=FALSE}
#full model
tapply(X=all.res[, "warns.full"]>0, INDEX=all.res[, c("r.effect")],
FUN=sum)
#warning codes: 
#363: unable to evaluate scaled gradient. Model failed to converge: degenerate  Hessian with 1 negative eigenvalues
#205: Model is nearly unidentifiable: large eigenvalue ratio - Rescale variables?
```



## Only models that converged are evaluated from here on:  

```{r include=FALSE}
all.res2=subset(all.res, warns.full==0)
```

### How many models converged, and a significant intercept?  
```{r echo=FALSE}

icpt.p.data <- all.res2 %>%
  group_by(test.per, r.effect) %>%
  summarise(mean.icpt.p = mean(icpt.p), 
            n.sign.icpt.p = length(icpt.p[icpt.p < 0.05]), 
            n.icpt.p = n.simus,#length(lrt.p.con), 
            proportion.sign.icpt.p = length(icpt.p[icpt.p < 0.05]) / n.simus)

icpt.p.data
```

#### Plotting the proportion of significant LRTs for the predictor variable condition ONLY based on models that converged and with a significant full-null model comparison

```{r echo=FALSE}

p.icpt.prop <- ggplot(data = icpt.p.data, aes(y = proportion.sign.icpt.p)) +
  geom_point(aes(x = as.factor(r.effect), y = proportion.sign.icpt.p)) +
  geom_boxplot(aes(x = as.factor(r.effect), y = proportion.sign.icpt.p, group = r.effect)) +
  ylim(0.0, 1) +
  geom_hline(yintercept = 0.8, colour = "blue", lwd = 1.05, lty = 3) +
  facet_wrap(~test.per) +
  ylab("Power") +
  xlab("Size of random intercept") +
  theme_bw()
p.icpt.prop

```

#### Plotting the intercepts

```{r echo=FALSE}

p.icpt.est <- ggplot(data = all.res2, aes(x = as.factor(r.effect), y = icpt)) +
  geom_jitter(alpha = 0.2, col = "grey") +
  geom_boxplot(aes(x = as.factor(r.effect), y = icpt, group = r.effect), alpha = 0.1, outlier.colour = "white") +
  facet_wrap(~test.per) +
  geom_hline(yintercept = 0, colour = "black", alpha = 0.5) +
  geom_hline(data = data.frame(test.per = "0.65"), aes(yintercept = coefs["(Intercept)"]), colour = "red", lwd = 1.05, lty = 2, alpha = 0.7) +
  geom_hline(data = data.frame(test.per = "0.55"), aes(yintercept = qlogis(0.55)), colour = "red", lwd = 1.05, lty = 2, alpha = 0.7) +
  ylab("Intercept") +
  xlab("Size of random intercept") +
  theme_bw()

p.int <- plot_grid(p.icpt.prop, p.icpt.est, labels = c("a", "b"))

ggsave(p.int, filename = "intercept_only_model.png", scale = 0.45, height = 7, width = 16)
```



#### Plotting the random intercept  

```{r echo=FALSE}
ggplot(data = all.res2, aes(x=as.factor(r.effect), y=re.sd))+
  geom_jitter( alpha=0.5, col="grey")+
  geom_boxplot(aes(x=as.factor(r.effect), y=re.sd, group=r.effect), alpha=0.1, outlier.colour="white")+
  facet_wrap(~test.per)

```



