---
title: "disjSyl_analysis_followup_script"
author: "Lou Haux, Hanna Schleihauf, Christoph Völter"
date: "8/12/2021"
output:   html_document:
    theme: united
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("dplyr")
library("ggplot2")
library("lme4")
library("effects")
library("car")
source("./functions/diagnostic_fcns.r")
source("./functions/glmm_stability.r")
source("./functions/boot_glmm.r")
source("./functions/drop1_para.r")
source("./functions/paired_data_functions.r")
```

**Load RData File**
```{r, include=FALSE}
load("./disjSyl_analysis_followup.RData")
```


# FOLLOW-UP EXPERIMENT

**Load and clean up data**
```{r, echo=FALSE}
# load data
xdata <- read.csv("./data/disjunctive_syllogism_data_experiment_followup.csv", sep = ';', na = c("NA")) 

# z-tranform covariates and create new variables for them
xdata$z.age <- as.vector(scale(xdata$age))
xdata$z.trial <- as.vector(scale(xdata$trial))

# transform character variables into factors
xdata$subj.id <- as.factor(xdata$subj.id)
xdata$sex <- as.factor(xdata$sex)
xdata$condition <- as.factor(xdata$condition) 
xdata$order <- as.factor(xdata$order) 
xdata$experience <- as.factor(xdata$experience)
xdata$other_pair <- as.factor(xdata$other_pair)

# dummy coding of conditon for the random slope part
xdata$condition.code = as.numeric(xdata$condition==levels(xdata$condition)[2])

str(xdata)
```


The dependent variable for this analysis is chimpanzees’ choice of the other pair

**Comparison of test condition: reveal-empty and control condition: reveal-baited **

*Fitting the full model*
```{r, echo=FALSE}
contr <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 10000000))

# fit full model:
full <- glmer(other_pair ~ condition + experience + z.age + sex + z.trial + order +
  (1 + condition.code + z.trial || subj.id),
data = xdata, family = binomial, control = contr
)
## in this model we did not include the correlation of random intercept and random slopes (indicated by ||), because it increases the chance that the model converges. 
```

*singular fit warnings*

#1 inspect results for random effects
```{r, echo=FALSE}
summary(full)$varcor
```

#2 exclude random slope of z.trials
```{r, echo=FALSE}
contr <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 10000000))

# fit full model:
full.1 <- glmer(other_pair ~ condition + experience + z.age + sex + z.trial + order +
  (1 + condition.code || subj.id),
data = xdata, family = binomial, control = contr
)
```

#3 exclude random slope of condition
```{r, echo=FALSE}
contr <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 10000000))

# fit full model:
full.2 <- glmer(other_pair ~ condition + experience + z.age + sex + z.trial + order +
  (1 | subj.id),
data = xdata, family = binomial, control = contr
)
```

# 4 check log-likelihood
```{r, echo=FALSE}
logLik(full)

logLik(full.1)

logLik(full.2)
# The log-likelihood reveals identical values
```

#5 check fixed effects
```{r, echo=FALSE}
round(summary(full)$coefficients, 3)

round(summary(full.1)$coefficients, 3)

round(summary(full.2)$coefficients, 3)
# There do not seem to be substantial differences
```

*Checking whether assumptions are fulfilled*
```{r, echo=FALSE}
# check for colliniarity
vif(full)

# checking random intercept and slopes
summary(full)$varcor
ranef.diagn.plot(full)

# checking model stability
m.stab.b <- glmm.model.stab(model.res = full, contr = contr, use = c("subj.id"))
m.stab.b$detailed$warnings
xx <- as.data.frame(round(m.stab.b$summary[, -1], 3))
m.stab.plot(round(m.stab.b$summary[, -1], 3))
```

*Fitting the null model*
```{r, echo=FALSE}
# fit null model:
null <- glmer(other_pair ~ experience + z.age + sex + z.trial + order +
  (1 + condition.code + z.trial || subj.id),
data = xdata, family = binomial, control = contr
)
```

*Comparing full and null model*
```{r, echo=FALSE}
round(anova(full, null, test = "Chisq"), 3)
```

*Comparing full and reduced models to get the effects of single predictors*
```{r, echo=FALSE}
tests <- drop1p(
  model.res = full, para = F, data = NULL, contr = contr,
  n.cores = c("all-1"), to.del = NULL
)
round(tests$drop1.res, 3)
## graphical depiction of the effects
plot(effect("condition", full))
plot(effect("experience", full))
plot(effect("z.age", full))
plot(effect("sex", full))
plot(effect("z.trial", full))
plot(effect("order", full))
```
*Looking at the estimates of the fixed effects*
```{r, echo=FALSE}
round(summary(full)$coefficients, 3)
```

*Calculating confidence intervals with 1000 bootstraps*
```{r, echo=FALSE}
# perform bootstraps for all predictors
boot.res <- boot.glmm.pred(
  model.res = full, excl.warnings = T,
  nboots = 1000, para = T, level = 0.95
)
round(boot.res$ci.estimates, 3)
xx <- as.data.frame(round(boot.res$ci.estimates, 3))
xx
m.stab.plot(round(boot.res$ci.estimates, 3))

# perform bootstraps for the main predictor of interest: condition (with the other predictors being at their average), this is necessary to plot the effect of condition
# dummy code and center all factors but condition
xdata$order.dummy <- as.numeric(xdata$order == levels(xdata$order)[2])
xdata$order.c <- xdata$order.dummy - mean(xdata$order.dummy)
xdata$sex.dummy <- as.numeric(xdata$sex == levels(xdata$sex)[2])
xdata$sex.c <- xdata$sex.dummy - mean(xdata$sex.dummy)
xdata$experience.dummy <- as.numeric(xdata$experience == levels(xdata$experience)[2])
xdata$experience.c <- xdata$experience.dummy - mean(xdata$experience.dummy)
# fit model with centered predictors
full.plot <- glmer(other_pair ~ condition + experience.c + z.age + sex.c + z.trial + order.c +
  (1 + condition + z.trial || subj.id),
data = xdata, family = binomial, control = contr
)
# perform bootstrap
boot.res.plot <- boot.glmm.pred(
  model.res = full.plot, excl.warnings = T,
  nboots = 1000, para = F, level = 0.95,
  use = c("condition")
)
round(boot.res.plot$ci.estimates, 3)
xx <- as.data.frame(round(boot.res.plot$ci.estimates, 3))
xx
m.stab.plot(round(boot.res.plot$ci.estimates, 3))

#estimates for the plot
boot.res.plot$ci.predicted
```
*Plots for the effect of test condition*
```{r, echo=FALSE, fig1, fig.height = 2, fig.width = 2}
#Plot for condition effect with model line
boot.plot = as.data.frame(boot.res.plot$ci.predicted)
rownames(boot.plot) = c("control", "test") #control = reveal-baited, test = reveal-empty
boot.plot$condition = rownames(boot.plot)
boot.plot$condition <- ordered(boot.plot$condition, levels = c("test", "control"))
xdata$fitted = as.numeric((xdata$other_pair)) - 1
theme_set( theme_gray()) 
xdata$condition <- ordered(xdata$condition, levels = c("test", "control"))
levels(xdata$condition)
table(xdata$other_pair , xdata$condition)
g <- ggplot(xdata, aes(x = condition , y = fitted, colour = condition))
g <- g +  geom_jitter(width = 0.3, height = 0.03, alpha = 0.3)
g <- g + geom_errorbar(aes(x=c("control"), 
                             ymin=c(boot.plot$lower.cl[boot.plot$condition == "control"]),
                             ymax=c(boot.plot$upper.cl[boot.plot$condition == "control"])),
                         width=.125, colour="#00CDCD",
                         position=position_dodge(0.9))
g <- g + geom_errorbar(aes(x=c("test"), 
                             ymin=c(boot.plot$lower.cl[boot.plot$condition == "test"]),
                             ymax=c(boot.plot$upper.cl[boot.plot$condition == "test"])),
                         width=.125, colour="#00CDCD",
                         position=position_dodge(0.9))
g <- g + geom_crossbar(data = boot.plot, aes(x=c("control", "test"), 
                                                 y = fitted, 
                                                 ymin= fitted,
                                                 ymax= fitted), 
                         width=0.5, colour= c("#00CDCD", "#00CDCD"))
g <- g +  scale_x_discrete(name = NULL, labels = c("reveal empty cup", "reveal baited cup"))
#g <- g + scale_y_continuous(labels = scales::percent_format(accuracy = 1))
g <- g +scale_y_continuous(name = "predicted probability of other-pair choices", breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1.0), labels = c("0.0", "0.2", "0.4", "0.6", "0.8", "1.0"))
g = g + theme(legend.position = "none",
                axis.text.x = element_text(size = 12, face=NULL, 
                                           margin = margin(t = 1, r = 0, b = 1, l = 0, 
                                                           unit = "mm")), 
                axis.text.y = element_text(size = 12, face=NULL,
                                           margin = margin(t = 0, r = 1, b = 0, l = 1, 
                                                           unit = "mm")),
                axis.title.x = element_text(size = 12, face=NULL,
                                            margin = margin(t = 1.5, r = 1, b = 0, l = 0, 
                                                            unit = "mm")),
                axis.title.y = element_text(size = 12, face=NULL,
                                            margin = margin(t = 1.5, r = 0, b = 0, l = 0, 
                                                            unit = "mm")))

g = g + scale_color_manual(values = c("#00CDCD", "#00CDCD"))  #, labels = c(" ", " "))

g

#ggsave(plot = g, filename = 'Follow.up.Experiment.Figure1.pdf', path = "./figures")
```


```{r,echo=FALSE, fig2, fig.height = 2.2, fig.width = 1.9}
#Plot for within-subject effects
sum.results = aggregate((as.numeric(xdata$other_pair[xdata$other_pair == "yes"]) - 1), 
                        by = list(xdata$subj.id[xdata$other_pair == "yes"], 
                                  xdata$condition[xdata$other_pair == "yes"]), 
                        FUN = sum)
colnames(sum.results) <- c("sub.id","condition", "other_pair")
sum.results$condition <- ordered(sum.results$condition, levels = c("test", "control"))
sum.results$other_pair <- sum.results$other_pair/16


#plot the result:
#pdf("./figures/Follow.up.Experiment.Figure2.pdf")
paired.data.plot(vector1=sum.results$other_pair[sum.results$condition == "test"], 
                      vector2=sum.results$other_pair[sum.results$condition == "control"], xlab="", 
                      ylab="proportion of other-pair choices", xnames=c("reveal empty cup", "reveal baited cup"), 
                      size.fac=c(1, NA, -1), ylim=(c(-0.05,1.05)), size.by="circles", edge.width=0.2, n1=NULL, 
                      n2=NULL, cex.lab=1, col.vec=NULL, lwd=1, bty="o", pch=NULL, lty=NULL, reset.par = TRUE,
                      add.y.axis=T, up.down.ltys=NULL, cex.axis=1, cex.axis.lab=1, main=NULL, cex.main=1)
#while (!is.null(dev.list()))  
#dev.off()
```


*Descriptive statistics*
```{r, echo=FALSE, count correct cups}
# For all chimps: control condition
xdata_control <- xdata %>% filter(condition=="control") %>% count(other_pair)
xdata_control

# For individual subjects: control condition
xdataind_control<- xdata %>% group_by(subj.id, other_pair) %>% filter(condition=="control") %>% count(other_pair)
xdataind_control<- pivot_wider(xdataind_control, names_from= other_pair, values_from= n) 
xdataind_control<- xdataind_control%>% select(subj.id, yes) %>% mutate(yes=yes/16)
xdataind_control_plot <- ggplot(data = xdataind_control, aes(y = yes, x = subj.id)) +  geom_boxplot() + ylab("Proportion of other-pair choices \n in the reveal-baited condition") + xlab("")  + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) + theme(axis.title = element_text(size = 15)) + theme(axis.text = element_text(size = 15))
xdataind_control_plot

# For all chimps: test condition
xdata_test <- xdata %>% filter(condition=="test") %>% count(other_pair)
xdata_test 

# For individual subjects: test condition
xdataind_test<- xdata %>% group_by(subj.id, other_pair) %>% filter(condition=="test") %>% count(other_pair)
xdataind_test<- pivot_wider(xdataind_test, names_from= other_pair, values_from= n) 
xdataind_test<- xdataind_test%>% select(subj.id, yes) %>% mutate(yes=yes/16)
xdataind_test_plot <- ggplot(data = xdataind_test, aes(y = yes, x = subj.id)) +  geom_boxplot() + ylab("Proportion of other-pair choices \n in the reveal-empty condition") + xlab("")  + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) + theme(axis.title = element_text(size = 15)) + theme(axis.text = element_text(size = 15))
xdataind_test_plot
```


```{r, echo=FALSE}
save.image("./disjSyl_analysis_followup.RData")
```
 
