---
title: " binomial GLMM"
author: ""
date: "31/05/2020"
output: 
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4)
library("lme4")
library("effects")
#source("/Users/HannaSchleihauf/Dropbox/Research/EVAClass/Functions/diagnostic_fcns.r")
```

**Load and clean up data**
```{r}
#load data
xdata = read.csv("data/dummy_data.csv",  na= c("NA"))
str(xdata)
#z-tranform covariates and create new variables for them
xdata$z.age = scale(xdata$age)
xdata$z.trial = scale(xdata$trial)
xdata$z.order = scale(xdata$order2)
#transform character variables into factors
xdata$subj.id = as.factor(xdata$subj.id)
xdata$condition = as.factor(xdata$condition)
xdata$order = as.factor(xdata$order)
xdata$sex = as.factor(xdata$sex)
#create dummy variables and centered dummy variables for the factors which are included in random slopes
xdata$condition.dummy=as.numeric(xdata$condition==levels(xdata$condition)[2])
xdata$condition.c = xdata$condition.dummy-mean(xdata$condition.dummy)
```

**Comparision of test condition and control condition**
*Fitting the full model*
```{r echo=FALSE}
  #this control term will be needed in case of convergance issues
  contr=glmerControl(optimizer="bobyqa", optCtrl=list(maxfun=10000000))

  # fit full model:  ####Sind wir hier sicher, dass wir keine interactionen drin haben wollen???
  full <- glmer(correct ~ condition + z.age + sex + z.trial + order + 
                  (1 + condition.c + z.trial || subj.id),
    data = xdata, family = binomial, control = contr
  )
  ##in this model we did not include the correlation of random intercept and random slopes (indicated by ||), because it increases the chance that the model converges. However, with the real data we will first try to fit the model with the correlations and only reduce it in case we have convergance issues or sigular fit messages. 
```

*Checking whether assumptions are fullfilled*
```{r}
  #checking absence of overdispersion
  overdisp.test(full) 
  #check for colliniarity
  xx=lm(correct ~ condition + z.age + sex + z.trial + order,
      data=xdata)
  library("car")
  vif(xx)
  #checking random intercept and slopes
  summary(full)$varcor
  ranef.diagn.plot(full)
  #checking model stability
  source("/Users/HannaSchleihauf/Dropbox/Research/EVAClass/Functions/glmm_stability.r")
  m.stab.b=glmm.model.stab(model.res=full, contr=contr, use=c("subj.id"))  
  m.stab.b$detailed$warnings
  xx=as.data.frame(round(m.stab.b$summary[, -1], 3))
  m.stab.plot(round(m.stab.b$summary[, -1], 3))
```

*Fitting the null model*
```{r}
  # fit null model:
    null <- glmer(correct ~ z.trial + order + (1 + condition.c + z.trial || subj.id),
    data = xdata, family = binomial, control = contr
  )
```

*Comparing full and null model*
```{r}
round(anova(full, null, test="Chisq"), 3)
```

*Comparing full and reduced models to get the effects of single predictors*
```{r}
##load function that drops one predictor at the time and compares it with the full model
source("/Users/HannaSchleihauf/Dropbox/Research/EVAClass/Functions/drop1_para.r") 
tests=drop1p(model.res=full, para=F, data=NULL, contr=contr, 
             n.cores=c("all-1"), to.del=NULL)
round(tests$drop1.res, 3)
##graphical dipiction of the effects
plot(effect("condition", full))
plot(effect("z.age", full))
plot(effect("sex", full))
plot(effect("z.trial", full))
plot(effect("order", full))
```

*Looking at the estimates of the fixed effects*
```{r}
round(summary(full)$coefficients, 3)
```

*Calculating confidence intervals with 1000 bootstraps*
```{r}
#load function to perform bootstraps
source("/Users/HannaSchleihauf/Dropbox/Research/EVAClass/Functions/boot_glmm.r")
#perform bootstraps for all predictors
boot.res=boot.glmm.pred(model.res=full, excl.warnings=T,
                        nboots=1000, para=F, level=0.95, 
                        use=c("condition", "z.age", "sex","z.trial", "order"))
round(boot.res$ci.estimates, 3)
xx=as.data.frame(round(boot.res$ci.estimates, 3))
xx
m.stab.plot(round(boot.res$ci.estimates, 3))

#perform bootstraps for the main predictor of interest: condition (with the other predictors being at their average), this is necessary to plot the effect of condition
#dummy code and center all factors but condition
xdata$order.dummy=as.numeric(xdata$order==levels(xdata$order)[2])
xdata$order.c = xdata$order.dummy-mean(xdata$order.dummy)
xdata$sex.dummy=as.numeric(xdata$sex==levels(xdata$sex)[2])
xdata$sex.c = xdata$sex.dummy-mean(xdata$sex.dummy)
#fit model with centered predictors
full.plot <- glmer(correct ~ condition + z.age + sex.c + z.trial + order.c + 
                  (1 + condition.c + z.trial || subj.id),
    data = xdata, family = binomial, control = contr
  )
#performm bootstrap
boot.res.plot=boot.glmm.pred(model.res=full.plot, excl.warnings=T,
                        nboots=1000, para=F, level=0.95, 
                        use=c("condition"))
round(boot.res.plot$ci.estimates, 3)
xx=as.data.frame(round(boot.res.plot$ci.estimates, 3))
xx
m.stab.plot(round(boot.res.plot$ci.estimates, 3))
```

**Chance level (0.5) comparision for the test condition**
*Subsetting the data set*
```{r}
xdata.test = subset(xdata, xdata$condition == "test")
```

*Fitting the intercept only model*
```{r}
chance.test <-  glmer(correct ~ 1 + 
                  (1 + z.trial || subj.id),
    data = xdata.test, family = binomial, control = contr
)

summary(chance.test)
##
##Lieber Christoph, du hast ja gesagt, dass der Intercept des Null Modells die Abweichung von 0.5 bei der mittleren Trial number (6.5) und in der Referenzkategorie von order (control-first) angibt, m체ssten wir dann nicht alle Pr채diktoren gecentered mit ins Modell nehmen, wie folgt: 
chance.test <-  glmer(correct ~ z.age + sex.c + z.trial + order.c + 
                  (1 + z.trial || subj.id),
    data = xdata.test, family = binomial, control = contr
)
```

##Hier bin ich mir leider gar nicht sicher, ob wir hier noch einen Vergleich rechnen w체rden, oder ob wir uns nur den p-wert des intercepts im Summary anschauen w체rden. 
