---
title: " binomial GLMM"
author: "Hanna Schleihauf, Lou Haux, Christoph Völter"
date: "31/05/2020"
output: 
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4)
library("lme4")
library("effects")
library("aod")
library("car")
source("./functions/diagnostic_fcns.r")
source("./functions/glmm_stability.r")
source("./functions/boot_glmm.r")
source("./functions/drop1_para.r")
```

**Load and clean up data**
```{r}
# load data
xdata <- read.csv("data/dummy_data.csv", na = c("NA")) 
str(xdata)
# z-tranform covariates and create new variables for them
xdata$z.age <- scale(xdata$age)
xdata$z.trial <- scale(xdata$trial)
# transform character variables into factors
xdata$subj.id <- as.factor(xdata$subj.id)
xdata$condition <- as.factor(xdata$condition) # only for Experiment 3 and 4
xdata$order <- as.factor(xdata$order) # only for Experiment 3 and 4
xdata$sex <- as.factor(xdata$sex)
# create dummy variables and centered dummy variables for the factors which are included in random slopes
xdata$condition.dummy <- as.numeric(xdata$condition == levels(xdata$condition)[2])
xdata$condition.c <- xdata$condition.dummy - mean(xdata$condition.dummy)
```

**For Experiment 1 & 2**
*Fitting the intercept-only model*
```{r}

# this control term will be needed in case of convergance issues
contr <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 10000000))

chance.test <- glmer(correct ~ 1 +
  (1 + z.trial || subj.id),
data = xdata, family = binomial, control = contr
)

summary(chance.test)
```


**For Experiment 3 and 4**

**Chance level (0.5) comparision for the test condition**
*Subsetting the data set*
```{r}
xdata.test <- subset(xdata, xdata$condition == "test")
```

*Fitting the intercept only model*
```{r}
chance.test <- glmer(correct ~ 1 +
  (1 + z.trial || subj.id),
data = xdata.test, family = binomial, control = contr
)

summary(chance.test)
```

**Comparision of test condition and control condition**
*Fitting the full model*
```{r echo=FALSE}
# this control term will be needed in case of convergance issues
contr <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 10000000))

# fit full model:
full <- glmer(correct ~ condition + z.age + sex + z.trial + order +
  (1 + condition.c + z.trial || subj.id),
data = xdata, family = binomial, control = contr
)
## in this model we did not include the correlation of random intercept and random slopes (indicated by ||), because it increases the chance that the model converges. However, with the real data we will first try to fit the model with the correlations and only reduce it in case we have convergance issues or sigular fit messages.
```

*Checking whether assumptions are fullfilled*
```{r}

# check for colliniarity
xx <- lm(correct ~ condition + z.age + sex + z.trial + order,
  data = xdata
)
vif(xx)
# checking random intercept and slopes
summary(full)$varcor
ranef.diagn.plot(full)
# checking model stability
m.stab.b <- glmm.model.stab(model.res = full, contr = contr, use = c("subj.id"))
m.stab.b$detailed$warnings
xx <- as.data.frame(round(m.stab.b$summary[, -1], 3))
m.stab.plot(round(m.stab.b$summary[, -1], 3))
```

*Fitting the null model*
```{r}
# fit null model:
null <- glmer(correct ~ z.trial + order + (1 + condition.c + z.trial || subj.id),
  data = xdata, family = binomial, control = contr
)
```

*Comparing full and null model*
```{r}
round(anova(full, null, test = "Chisq"), 3)
```

*Comparing full and reduced models to get the effects of single predictors*
```{r}

tests <- drop1p(
  model.res = full, para = F, data = NULL, contr = contr,
  n.cores = c("all-1"), to.del = NULL
)
round(tests$drop1.res, 3)
## graphical dipiction of the effects
plot(effect("condition", full))
plot(effect("z.age", full))
plot(effect("sex", full))
plot(effect("z.trial", full))
plot(effect("order", full))
```

*Looking at the estimates of the fixed effects*
```{r}
round(summary(full)$coefficients, 3)
```

*Calculating confidence intervals with 1000 bootstraps*
```{r}

# perform bootstraps for all predictors
boot.res <- boot.glmm.pred(
  model.res = full, excl.warnings = T,
  nboots = 1000, para = T, level = 0.95
)
round(boot.res$ci.estimates, 3)
xx <- as.data.frame(round(boot.res$ci.estimates, 3))
xx
m.stab.plot(round(boot.res$ci.estimates, 3))

# perform bootstraps for the main predictor of interest: condition (with the other predictors being at their average), this is necessary to plot the effect of condition
# dummy code and center all factors but condition
xdata$order.dummy <- as.numeric(xdata$order == levels(xdata$order)[2])
xdata$order.c <- xdata$order.dummy - mean(xdata$order.dummy)
xdata$sex.dummy <- as.numeric(xdata$sex == levels(xdata$sex)[2])
xdata$sex.c <- xdata$sex.dummy - mean(xdata$sex.dummy)
# fit model with centered predictors
full.plot <- glmer(correct ~ condition + z.age + sex.c + z.trial + order.c +
  (1 + condition.c + z.trial || subj.id),
data = xdata, family = binomial, control = contr
)
# performm bootstrap
boot.res.plot <- boot.glmm.pred(
  model.res = full.plot, excl.warnings = T,
  nboots = 1000, para = T, level = 0.95,
  use = c("condition")
)
round(boot.res.plot$ci.estimates, 3)
xx <- as.data.frame(round(boot.res.plot$ci.estimates, 3))
xx
m.stab.plot(round(boot.res.plot$ci.estimates, 3))
```



 
