---
title: "disjSyl_analysis_followup_script"
author: "Lou Haux, Hanna Schleihauf, Christoph Völter"
date: "8/30/2021"
output:   html_document:
    theme: united
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("dplyr")
library("ggplot2")
library("gghalves")
library("lme4")
library("effects")
library("car")
source("./functions/diagnostic_fcns.r")
source("./functions/glmm_stability.r")
source("./functions/boot_glmm.r")
source("./functions/drop1_para.r")
source("./functions/paired_data_functions.r")
```

**Load RData File**
```{r, include=FALSE}
load("./disjSyl_analysis_followup.RData")
```


# FOLLOW-UP EXPERIMENT

**Load and clean up data**
```{r, echo=FALSE}
# load data
xdata <- read.csv("./data/disjunctive_syllogism_data_experiment_followup.csv", sep = ';', na = c("NA")) 

# z-tranform covariates and create new variables for them
xdata$z.age <- as.vector(scale(xdata$age))
xdata$z.trial <- as.vector(scale(xdata$trial))

# transform character variables into factors
xdata$subj.id <- as.factor(xdata$subj.id)
xdata$sex <- as.factor(xdata$sex)
xdata$condition <- as.factor(xdata$condition) 
xdata$order <- as.factor(xdata$order) 
xdata$experience <- as.factor(xdata$experience)
xdata$other_pair <- as.factor(xdata$other_pair)

# dummy coding of conditon for the random slope part
xdata$condition.code = as.numeric(xdata$condition==levels(xdata$condition)[2])

str(xdata)
```


The dependent variable for this analysis is chimpanzees’ choice of the other pair

**Comparison of test condition: reveal-empty and control condition: reveal-baited **

*Fitting the full model*
```{r, echo=FALSE}
contr <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 10000000))

# fit full model:
full <- glmer(other_pair ~ condition + z.age + sex + z.trial + order +
  (1 + condition.code + z.trial | subj.id),
data = xdata, family = binomial, control = contr
)
## in this model we did include the correlation of random intercept and random slopes 
```

*singular fit warnings*

#1 inspect results for random effects
```{r, echo=FALSE}
summary(full)$varcor
```

#2 exclude correlations
```{r, echo=FALSE}
contr <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 10000000))

# fit full model:
full.1 <- glmer(other_pair ~ condition + z.age + sex + z.trial + order +
  (1 + condition.code + z.trial || subj.id),
data = xdata, family = binomial, control = contr
)

summary(full.1)$varcor
## in this model we did not include the correlation of random intercept and random slopes 
```

#3 exclude random slope of z.trials
```{r, echo=FALSE}
contr <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 10000000))

# fit full model:
full.2 <- glmer(other_pair ~ condition + z.age + sex + z.trial + order +
  (1 + condition.code || subj.id),
data = xdata, family = binomial, control = contr
)

summary(full.2)$varcor
## in this model we did not include the correlation of random intercept and random slopes 
```

#4 exclude random slope of condition
```{r, echo=FALSE}
contr <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 10000000))

# fit full model:
full.3 <- glmer(other_pair ~ condition + z.age + sex + z.trial + order +
  (1 | subj.id),
data = xdata, family = binomial, control = contr
)

summary(full.3)$varcor
```

# 5 check log-likelihood
```{r, echo=FALSE}
logLik(full)

logLik(full.1)

logLik(full.2)

logLik(full.3)
# The log-likelihood reveals identical values
```

#6 check fixed effects
```{r, echo=FALSE}
round(summary(full)$coefficients, 3)

round(summary(full.1)$coefficients, 3)

round(summary(full.2)$coefficients, 3)

round(summary(full.3)$coefficients, 3)

# There do not seem to be substantial differences
```

*Checking whether assumptions are fulfilled*
```{r, echo=FALSE}
# check for colliniarity
vif(full)

# checking random intercept and slopes
summary(full)$varcor
ranef.diagn.plot(full)

# checking model stability
m.stab.b <- glmm.model.stab(model.res = full, contr = contr, use = c("subj.id"))
m.stab.b$detailed$warnings
xx <- as.data.frame(round(m.stab.b$summary[, -1], 3))
m.stab.plot(round(m.stab.b$summary[, -1], 3))
```

*Fitting the null model*
```{r, echo=FALSE}
# fit null model:
null <- glmer(other_pair ~ z.trial + order +
  (1 + condition.code + z.trial | subj.id),
data = xdata, family = binomial, control = contr
)
```

*Comparing full and null model*
```{r, echo=FALSE}
round(anova(full, null, test = "Chisq"), 3)
```

*Comparing full and reduced models to get the effects of single predictors*
```{r, echo=FALSE}
tests <- drop1p(
  model.res = full, para = F, data = NULL, contr = contr,
  n.cores = c("all-1"), to.del = NULL
)
round(tests$drop1.res, 3)
## graphical depiction of the effects
plot(effect("condition", full))
plot(effect("z.age", full))
plot(effect("sex", full))
plot(effect("z.trial", full))
plot(effect("order", full))
```
*Looking at the estimates of the fixed effects*
```{r, echo=FALSE}
round(summary(full)$coefficients, 3)
```

*Calculating confidence intervals with 1000 bootstraps*
```{r, echo=FALSE}
# perform bootstraps for all predictors
boot.res <- boot.glmm.pred(
  model.res = full, excl.warnings = T,
  nboots = 1000, para = T, level = 0.95
)
round(boot.res$ci.estimates, 3)
xx <- as.data.frame(round(boot.res$ci.estimates, 3))
xx
m.stab.plot(round(boot.res$ci.estimates, 3))

# perform bootstraps for the main predictor of interest: condition (with the other predictors being at their average), this is necessary to plot the effect of condition
# dummy code and center all factors but condition
xdata$order.dummy <- as.numeric(xdata$order == levels(xdata$order)[2])
xdata$order.c <- xdata$order.dummy - mean(xdata$order.dummy)
xdata$sex.dummy <- as.numeric(xdata$sex == levels(xdata$sex)[2])
xdata$sex.c <- xdata$sex.dummy - mean(xdata$sex.dummy)

# fit model with centered predictors
full.plot <- glmer(other_pair ~ condition + z.age + sex.c + z.trial + order.c +
  (1 + condition + z.trial || subj.id),
data = xdata, family = binomial, control = contr
)
# perform bootstrap
boot.res.plot <- boot.glmm.pred(
  model.res = full.plot, excl.warnings = T,
  nboots = 1000, para = F, level = 0.95,
  use = c("condition")
)
round(boot.res.plot$ci.estimates, 3)
xx_plot <- as.data.frame(round(boot.res.plot$ci.estimates, 3))
xx_plot
m.stab.plot(round(boot.res.plot$ci.estimates, 3))

#estimates for the plot
boot.res.plot$ci.predicted
```

*Plots for the effect of test condition*

```{r}

library(gghalves)
library(ggthemes)
library(cowplot)


xdata.agg <- xdata %>%
  mutate(other_pair2=as.numeric(other_pair)-1)%>%
  group_by(subj.id, condition) %>%
  summarise(mean.resp = mean(other_pair2)) %>%
  ungroup()

xdata.agg$condition2 <- jitter(as.numeric(as.factor(xdata.agg$condition), amount = .0001))


followup_plot <- ggplot(data = xdata.agg, aes(x = condition, y = mean.resp)) +
  #geom_hline(yintercept = 2/3, lty = 2, col = "black", alpha=0.3) +
  geom_point(data = xdata.agg %>% filter(condition == "control"), aes(x = condition2), color = "dodgerblue", size = 1.5, alpha = .5) +
  geom_point(data = xdata.agg %>% filter(condition == "test"), aes(x = condition2), color = "darkorange", size = 1.5, alpha = .5, ) +
  geom_line(aes(x = condition2, group = subj.id), color = "gray", lty = 1, alpha = .3) +
  #geom_half_violin(data = xdata3.agg %>% filter(condition == "control"), aes(x = condition2, y = mean.resp), position = position_nudge(x = -.5), side = "l", width = .2, fill = "dodgerblue", alpha = .5) +
  geom_half_boxplot(data = xdata.agg %>% filter(condition == "control"), aes(x = condition2, y = mean.resp), position = position_nudge(x = -.7),  width = .2, side = "r",outlier.shape = NA, center = TRUE, errorbar.draw = FALSE, fill = "dodgerblue", alpha = .5) +
  #geom_half_violin(data = xdata3.agg %>% filter(condition == "test"), aes(x = condition2, y = mean.resp), position = position_nudge(x = 0.5), side = "r", width = .2, fill = "darkorange", alpha = .5) +
  geom_half_boxplot(data = xdata.agg %>% filter(condition == "test"), aes(x = condition2, y = mean.resp), position = position_nudge(x = 0.5), width = .2, side = "r",outlier.shape = NA, center = TRUE, errorbar.draw = FALSE,  fill = "darkorange", alpha = .5) +
    geom_errorbar(data = boot.res.plot$ci.predicted %>% filter(condition == "test"), aes(x = 2.3, y = fitted, ymin = lower.cl, ymax = upper.cl), color = "darkorange", width = 0.1, size = 0.6) +
  geom_errorbar(data = boot.res.plot$ci.predicted %>% filter(condition == "control"), aes(x = 0.65, y = fitted, ymin = lower.cl, ymax = upper.cl), color = "dodgerblue", width = 0.1, size = 0.6) +
  geom_point(data = boot.res.plot$ci.predicted %>% filter(condition == "test"), aes(x = 2.3, y = fitted), color = "darkorange", size = 1.5) +
  geom_point(data = boot.res.plot$ci.predicted %>% filter(condition == "control"), aes(x = 0.65, y = fitted), color = "dodgerblue", size = 1.5) +


  # Define additional settings
  xlab("Condition") +
  ylab("Proportion of other-pair choices") +
  scale_x_continuous(breaks = c(0.75, 2.25), labels = c("Reveal baited", "Reveal empty"), limits = c(0,3)) +
  ylim(0, 1) +
  theme_classic()

followup_plot

ggsave(followup_plot, filename = "figures/Follow_up_Experiment_Figure3.png", width = 5, height = 5, scale = 0.55)


```


*Descriptive statistics*
```{r, echo=FALSE, count correct cups}
# For all chimps: control condition
xdata_control <- xdata %>% filter(condition=="control") %>% count(other_pair)
xdata_control

# For individual subjects: control condition
xdataind_control<- xdata %>% group_by(subj.id, other_pair) %>% filter(condition=="control") %>% count(other_pair)
xdataind_control<- pivot_wider(xdataind_control, names_from= other_pair, values_from= n) 
xdataind_control<- xdataind_control%>% select(subj.id, yes) %>% mutate(yes=yes/16)
xdataind_control_plot <- ggplot(data = xdataind_control, aes(y = yes, x = subj.id)) +  geom_boxplot() + ylab("Proportion of other-pair choices \n in the reveal-baited condition") + xlab("")  + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) + theme(axis.title = element_text(size = 15)) + theme(axis.text = element_text(size = 15))
xdataind_control_plot

# For all chimps: test condition
xdata_test <- xdata %>% filter(condition=="test") %>% count(other_pair)
xdata_test 

# For individual subjects: test condition
xdataind_test<- xdata %>% group_by(subj.id, other_pair) %>% filter(condition=="test") %>% count(other_pair)
xdataind_test<- pivot_wider(xdataind_test, names_from= other_pair, values_from= n) 
xdataind_test<- xdataind_test%>% select(subj.id, yes) %>% mutate(yes=yes/16)
xdataind_test_plot <- ggplot(data = xdataind_test, aes(y = yes, x = subj.id)) +  geom_boxplot() + ylab("Proportion of other-pair choices \n in the reveal-empty condition") + xlab("")  + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) + theme(axis.title = element_text(size = 15)) + theme(axis.text = element_text(size = 15))
xdataind_test_plot
```




*secondary analysis*

# To test the effect of previous experience

*Fitting the full model*
```{r, echo=FALSE}
contr <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 10000000))

# fit full_with_exp model:
full_with_exp <- glmer(other_pair ~ condition + z.age + sex + experience + z.trial + order +
  (1 + condition.code + z.trial | subj.id),
data = xdata, family = binomial, control = contr
)
## in this model we did include the correlation of random intercept and random slopes 
```

*Fitting the null_with_exp model*
```{r, echo=FALSE}
# fit null_with_exp model:
null_with_exp <- glmer(other_pair ~ experience + z.trial + order +
  (1 + condition.code + z.trial | subj.id),
data = xdata, family = binomial, control = contr
)
```

*Comparing full_with_exp and null_with_exp model*
```{r, echo=FALSE}
round(anova(full_with_exp, null_with_exp, test = "Chisq"), 3)
```

*Comparing full_with_exp and reduced models to get the effects of single predictors*
```{r, echo=FALSE}
tests <- drop1p(
  model.res = full_with_exp, para = F, data =  NULL, contr = contr,
  n.cores = c("all-1"), to.del = NULL
)
round(tests$drop1.res, 3)
## graphical depiction of the effects
plot(effect("condition", full_with_exp))
plot(effect("experience", full_with_exp))
plot(effect("z.age", full_with_exp))
plot(effect("sex", full_with_exp))
plot(effect("z.trial", full_with_exp))
plot(effect("order", full_with_exp))
```
*Looking at the estimates of the fixed effects*
```{r, echo=FALSE}
round(summary(full_with_exp)$coefficients, 3)
```

*Calculating confidence intervals with 1000 bootstraps*
```{r, echo=FALSE}
# perform bootstraps for all predictors
boot.res.exp <- boot.glmm.pred(
  model.res = full_with_exp, excl.warnings = T,
  nboots = 1000, para = T, level = 0.95
)
round(boot.res.exp$ci.estimates, 3)
xx_with_exp <- as.data.frame(round(boot.res.exp$ci.estimates, 3))
xx_with_exp
m.stab.plot(round(boot.res.exp$ci.estimates, 3))

# perform bootstraps for the main predictor of interest: condition (with the other predictors being at their average), this is necessary to plot the effect of condition
# dummy code and center all factors but condition
xdata$order.dummy <- as.numeric(xdata$order == levels(xdata$order)[2])
xdata$order.c <- xdata$order.dummy - mean(xdata$order.dummy)
xdata$sex.dummy <- as.numeric(xdata$sex == levels(xdata$sex)[2])
xdata$sex.c <- xdata$sex.dummy - mean(xdata$sex.dummy)
xdata$experience.dummy <- as.numeric(xdata$experience == levels(xdata$experience)[2])
xdata$experience.c <- xdata$experience.dummy - mean(xdata$experience.dummy)
# fit model with centered predictors
full_with_exp.plot <- glmer(other_pair ~ condition + experience.c + z.age + sex.c + z.trial + order.c +
  (1 + condition + z.trial || subj.id),
data = xdata, family = binomial, control = contr
)
# perform bootstrap
boot.res.exp.plot <- boot.glmm.pred(
  model.res = full_with_exp.plot, excl.warnings = T,
  nboots = 1000, para = F, level = 0.95,
  use = c("condition")
)
round(boot.res.exp.plot$ci.estimates, 3)
xx_with_exp_plot <- as.data.frame(round(boot.res.exp.plot$ci.estimates, 3))
xx_with_exp_plot
m.stab.plot(round(boot.res.exp.plot$ci.estimates, 3))

#estimates for the plot
boot.res.exp.plot$ci.predicted
```


```{r, echo=FALSE}
save.image("./disjSyl_analysis_followup.RData")
```
 
